---
title: "NYC Property Sales Analysis"
author: "Matthew Scott"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

# Load required libraries
library(tidyverse)
library(caret)
library(readr)
library(ggplot2)
library(dplyr)
library(ggcorrplot)
library(xgboost)
library(lubridate)
library(gridExtra)
library(randomForest)
library(doParallel)
library(lightgbm)

# Source the main analysis script
source("NYCPropertySales_Advanced.R")
```

# Executive Summary

This report presents a comprehensive analysis of New York City property sales data, focusing on residential properties across different boroughs. We employ various machine learning models to predict property prices and analyze market trends.

# Data Overview

The analysis utilizes two primary datasets:
1. NYC Property Sales data (`nyc-rolling-sales.csv`)
2. NYC Building Codes (`NYC_Codes.csv`)

## Data Summary

```{r data-summary}
# Create summary statistics
summary_stats <- nyc_rolling_sales %>%
  summarise(
    Total_Properties = n(),
    Avg_Price = mean(`SALE PRICE`, na.rm = TRUE),
    Median_Price = median(`SALE PRICE`, na.rm = TRUE),
    Avg_SqFt = mean(`GROSS SQUARE FEET`, na.rm = TRUE),
    Median_SqFt = median(`GROSS SQUARE FEET`, na.rm = TRUE)
  )

# Display formatted summary
knitr::kable(summary_stats, 
             format.args = list(big.mark = ","),
             caption = "Summary Statistics")

# Borough-wise summary
borough_summary <- nyc_rolling_sales %>%
  group_by(borough_name) %>%
  summarise(
    Properties = n(),
    Avg_Price = mean(`SALE PRICE`, na.rm = TRUE),
    Median_Price = median(`SALE PRICE`, na.rm = TRUE)
  )

knitr::kable(borough_summary, 
             format.args = list(big.mark = ","),
             caption = "Borough-wise Summary")
```

# Exploratory Data Analysis

## Price Distributions

```{r price-distributions, fig.height=8}
# Create price distribution plots
p1 <- ggplot(nyc_rolling_sales, aes(x = `SALE PRICE`)) +
  geom_histogram(binwidth = 50000, fill = "steelblue", color = "black") +
  labs(title = "Original Sale Price Distribution", x = "Sale Price", y = "Count") +
  scale_x_continuous(labels = scales::comma)

p2 <- ggplot(nyc_rolling_sales, aes(x = log_sale_price)) +
  geom_histogram(binwidth = 0.1, fill = "steelblue", color = "black") +
  labs(title = "Log-Transformed Sale Price Distribution", x = "Log Sale Price", y = "Count")

p3 <- ggplot(nyc_rolling_sales, aes(x = `GROSS SQUARE FEET`)) +
  geom_histogram(binwidth = 100, fill = "steelblue", color = "black") +
  labs(title = "Sq Ft Distribution", x = "SqFt", y = "Count") +
  scale_x_continuous(labels = scales::comma)

p4 <- ggplot(nyc_rolling_sales, aes(x = log_gross_sqft)) +
  geom_histogram(binwidth = 0.1, fill = "steelblue", color = "black") +
  labs(title = "Log Sq Ft Distribution", x = "Log SqFt", y = "Count")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## Geographic Analysis

```{r geographic-analysis}
# Create borough-wise price distribution
ggplot(nyc_rolling_sales, aes(x = borough_name, y = `SALE PRICE`)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(title = "Price Distribution by Borough",
       x = "Borough",
       y = "Sale Price ($)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Create neighborhood analysis plot
p5 <- ggplot(nyc_rolling_sales, aes(x = `SALE PRICE`, y = NEIGHBORHOOD, color = borough_name)) +
  geom_boxplot(alpha = 0.6) +
  facet_wrap(~borough_name, scales = "free") +
  labs(
    title = "Sale Price Distribution by Neighborhood and Borough",
    y = "Neighborhood",
    x = "Sale Price ($)"
  ) +
  scale_x_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top")

print(p5)
```

# Model Performance Analysis

## Model Comparison

```{r model-comparison}
# Create comparison visualizations
test_results_comprehensive <- data.frame(
  actual = exp(y_test),
  xgb_predicted = xgb_predictions_exp,
  rf_predicted = rf_predictions_exp,
  lgb_predicted = lgb_predictions_exp,
  borough = X_test$borough_name,
  desc = X_test$Description
)

# Borough-wise comparison
p7 <- test_results_comprehensive %>%
  pivot_longer(
    cols = c(xgb_predicted, rf_predicted, lgb_predicted),
    names_to = "model",
    values_to = "predicted"
  ) %>%
  ggplot(aes(x = actual, y = predicted, color = model)) +
  geom_point(alpha = 0.25) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_wrap(~borough, scales = "free") +
  labs(
    title = "Model Comparison: Predicted vs Actual Values by Borough",
    x = "Actual Sale Price ($)",
    y = "Predicted Sale Price ($)"
  ) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

print(p7)

# Model-wise comparison
p8 <- test_results_comprehensive %>%
  pivot_longer(
    cols = c(xgb_predicted, rf_predicted, lgb_predicted),
    names_to = "model",
    values_to = "predicted"
  ) %>%
  ggplot(aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.25) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Model Comparison: Predicted vs Actual Values by Model Type",
    x = "Actual Sale Price ($)",
    y = "Predicted Sale Price ($)"
  ) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

print(p8)
```

## Error Analysis

```{r error-analysis}
# Create error distribution plot
error_data <- test_results_comprehensive %>%
  mutate(
    xgb_error = abs(actual - xgb_predicted)/actual * 100,
    rf_error = abs(actual - rf_predicted)/actual * 100,
    lgb_error = abs(actual - lgb_predicted)/actual * 100
  ) %>%
  pivot_longer(
    cols = c(xgb_error, rf_error, lgb_error),
    names_to = "model",
    values_to = "error"
  ) %>%
  mutate(model = gsub("_error", "", model))

p9 <- ggplot(error_data, aes(x = error, fill = model)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Error Distribution by Model",
    x = "Percentage Error",
    y = "Density",
    fill = "Model"
  ) +
  theme_minimal() +
  xlim(0, 100)

print(p9)

# Calculate error metrics
error_metrics <- test_results_comprehensive %>%
  summarise(
    XGBoost_MAPE = mean(abs(actual - xgb_predicted)/actual * 100),
    RF_MAPE = mean(abs(actual - rf_predicted)/actual * 100),
    LightGBM_MAPE = mean(abs(actual - lgb_predicted)/actual * 100)
  )

knitr::kable(error_metrics, 
             caption = "Mean Absolute Percentage Error by Model",
             digits = 2)
```

# Conclusions

1. **Price Distribution**: Property prices in NYC show significant variation across boroughs and neighborhoods, with Manhattan consistently showing higher median prices.

2. **Size-Price Relationship**: There is a strong correlation between property size and price, though this relationship varies by borough.

3. **Model Performance**:
   - XGBoost shows robust performance across different price ranges
   - Random Forest provides consistent predictions
   - LightGBM demonstrates competitive results, particularly in mid-range properties

# Recommendations

1. **For Buyers**: Consider properties in emerging neighborhoods that show stable price trends but haven't yet reached peak valuations.

2. **For Sellers**: Use the model predictions as a baseline for pricing strategy, considering the specific characteristics of the property and neighborhood.

3. **For Future Analysis**: 
   - Incorporate additional features such as proximity to amenities
   - Develop borough-specific models for more accurate predictions
   - Include temporal analysis to capture market trends 